Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-03 demystify.tex
updates

In your statement, you have the opportunity to connect directly with the adcom and make your application stand out and unique. Your SOP can convince the adcom that you would \emph{fit} the program, even if you don't have very strong research experience.
A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

There are various guides on writing SOP , and \href{https://cs-sop.org/}{many example statements} are available. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidences. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..).

You should also talk about things that adcom members might not know about and can help make you \textbf{stand out} in the application pool of  thousands of applicants, e.g., you are member of a minority or LGBTQ+ group (\autoref{sec:urm}), your personal Github project with 1K+ stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Finally, you should get your SOP reviewed by your LoR writers, professors (especially those who have served in adcom), or even graduate PhD students (because they have been through this process).

\begin{commentbox}[Vu:]
  I often read SOP and LORs first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something really standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or  adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to wrong schools or mixing
  facts (e.g., talking about school X but mentioned about working with
  profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
  should.
\end{commentbox}

@@

@@

\textbf{Advice}: If you think your school is the top in your country but might not be well-known internationally, you should talk about it in your SOP (and also ask your LOR writer to do that). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\textbf{Advice}: As with school reputation, you and your LoR writers can mention about the grading system of your university if you think that is helpful for adcom to evaluate.
Note that while having good grades might not help much,
having bad grades (e.g., < 3.0) will be \red{red flag} (unless your LORs or
statement give a proper explanation). This is especially true if you
have bad grades in relevant courses, e.g., Math and CS.

@@

\paragraph{Open House} Most schools have \emph{Open House} or \emph{Visit Day} events, which are a great resource to learn about the school, department, faculty, research, living, etc.
\textbf{Advice}: Even if you can't come in person, you should attend virtually and meet with individual faculty. During the event, you get a chance to learn more about the program, talk to individual faculty and current students.  Take notes of faculty who make you excited, count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, funding situation etc.  Ask about anything you want to determine that they deserve \emph{you}.

@@

\paragraph{What else?} Make decision, accept, reject, or defer the offers  (\autoref{sec:accept-postpone-decline}). Ask to meet with potential advisers (e.g., through Open House or separately) and even their students. Ask about computer equipment and software, office space, and other resources; in many case these will be provided for free by your adviser or department (\autoref{sec:buying-equipment}).

@@

Understandably students get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your adviser first.  They might have funding to buy you computers and other equipments (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide some budget or computer equipments to incoming PhD students (e.g., a laptop).

\section{Can I negotiate my PhD offer?}\label{sec:negotiate}
For stipend, unlikely, as TA is set by the university and while RA is set by the adviser, it is often based on the department's standard, and it also would be unfair to other students if you get more. However, your stipend is often increased every year (automatically by a small amount) and also after you pass your qualifying exam.

For start date or TA assignment (e.g., TA'ing a specific course), it is uncommon but you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely non-negotiable.  However, you can ask for things such as computer equipments (\autoref{sec:buying-equipment}), books, (and sometimes even furniture) and conference travel budget.

>>>>

In your statement, you have the opportunity to connect directly with us and make your application stand out and unique. Your SOP can convince adcom members that you would \emph{fit} the program, even if you don't have very strong research experience.
A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

There are various guides on writing SOP , and \href{https://cs-sop.org/}{many example statements} are available. In short, focus on your research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidences. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..).

You should talk about things that adcom members might not know about and can help make you \textbf{stand out} in the application pool of  thousands of applicants, e.g., you are member of a minority or LGBTQ+ group (\autoref{sec:urm}), your personal Github project with hundreds of stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Finally, you should get your SOP reviewed by your LoR writers, professors (especially those who have served in adcom), or even postdocs or PhD students (because they have been through this process).

\begin{commentbox}[Vu:]
  I often read LORs and SOP first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something really standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or  adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to wrong schools or mixing
  facts (e.g., talking about school X but mentioned about working with
  profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
  should.
\end{commentbox}

@@

This section discusses various research experiences that can strengthen your application.
\autoref{sec:research-opportunities} provides more information on how to find research opportunities (e.g., during your undergrad study).

@@

\textbf{Advice}: If your school is the top in your country but you think it might not be well-known internationally, you should talk about it in your SOP (and also ask your LOR writer to do that). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\textbf{Advice}: As with school reputation, you and your LoR writers can mention the grading system of your university if you think that is helpful for adcom to evaluate.
Note that while having good grades might not help much,
having bad grades (e.g., < 3.0) will be \red{red flag} (unless your LORs or
statement give a proper explanation). This is especially true if you
have bad grades in relevant courses, e.g., Math and CS.

@@

\paragraph{Open House} Most schools have \emph{Open House} or \emph{Visit Day} events, which are a great resource to learn about the school, department, faculty, research, living, etc.
\textbf{Advice}: Even if you can't come in person, you should attend virtually and meet with individual faculty. During the event, you get a chance to learn more about the program, and talk to individual faculty and current students.  Take notes of faculty who make you excited, and count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, and funding situation.  Ask about anything you want to determine that they deserve \emph{you}.

@@

\paragraph{What else?} Make a decision, accept, reject, or defer the offers  (\autoref{sec:accept-postpone-decline}). Ask to meet with potential advisers (e.g., through Open House or separately) and even their students. Ask about computer equipment and software, office space, and other resources; in many cases these will be provided for free by your adviser or department (\autoref{sec:buying-equipment}).

@@

Understandably students get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your adviser first.  They might have funding to buy you computers and other equipment (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide some budget or computer equipment to incoming PhD students (e.g., a laptop).

\section{Can I negotiate my PhD offer (e.g., if I have multiple offers)?}\label{sec:negotiate}

For stipend, unlikely for TA, which is set by the university. Moreover, while RA is set by the adviser, it is often based on the department's standard, and it also would be unfair to other students if you get more.  However, your stipend is often increased every year (automatically by a small amount) and also after you pass your qualifier exams.

For start date or TA assignment (e.g., TA'ing a specific course), it is uncommon but you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely non-negotiable.  However, you can ask for things such as computer equipment (\autoref{sec:buying-equipment}), books, (and sometimes even furniture), and a conference travel budget.
